Implicit Neural Representation for Potential Field Geophysics

### 2023-10-10 discussion.tex
Further fix cross references

Super-resolution for geophysics is still in its infancy and requires further efforts towards improving reliability before wide-spread adoption in practice.
This section proposes four avenues for future research:
\begin{enumerate}
    \item{} Exploring the outcomes from the use of SR in the task of geological mapping
    \item{} Understanding the impact of choices in constructing a model and dataset for SR geophysics for a range of geological domains,
    \item{} Identifying an improved image quality assessment for gridded potential field data.
    \item{} Investigating the use of a physics informed approach for SR and implicit representation, and,
\end{enumerate}

\section{SR impact on interpretation}
The super-resolution outputs from the methods presented in this thesis have improved structural accuracy over the low-resolution grids.
This is likely to have a positive impact on the detection of features such as lineaments and unit boundaries for geological interpretation from magnetic data.
However, despite the reduction of introduced structural inaccuracies with the use of open access training data in Cref{ch:paper2}, it is likely that some features with specific structural geometries will still be incorrectly super-resolved.
A domain expert may be able to recognise structures that are inconsistent using auxiliary information and the regional geological context, but machine processed super-resolution grids will use these data as provided.
In order to improve the adoption of SR for geological interpretation from potential field data, a rigorous study of the impact of SR on interpretation outcomes is required.
This could be a geological mapping task, such as the reinterpretation of regional geology using super-resolution upscaled regional magnetic TMI grids.
Comparison could then be made for the accuracy and benefit of a super-resolution method compared to the existing low-resolution interpretation.
However, such analysis will be difficult to perform with respect to different interpreters' variability in pattern recognition, especially in the context of their confidence in the reliability of the super-resolution product.
Previously, \textcite{sivarajahIdentifyingEffectiveInterpretation2013} reported significant variability between individuals when detecting geological targets from magnetic geophysics data.
A more reliable, repeatable and quantifiable approach would use automated pattern recognition methods developed for geophysical features.
For example, the automated lineament detection technique of \textcite{holdenAutomaticIdentificationResponses2011,holdenAutomatedAnalysisRegional2008} could be applied to low-resolution and SR grids, to compare their delineated features.

\section{SR model construction}
Several models were implemented throughout cref{ch:paper1, ch:paper2}, namely RDN, ESRGAN, and LTE\@.
% However, the most important choice in developing a performant super-resolution neural network for real-world aeromagnetic survey data was the choice of training data.
Cref{paper1} was prepared prior the release of a large and varied synthetic geology training dataset \parencite{jessellNoddyverseMassiveData2022}.
It was seen in cref{paper2} that this data improved model performance for a range of features by using transfer learning for pre-training.
% This transfer learning approach for training data is also found to be useful for RDN and ESRGAN\@.
Here, the use of synthetic and real-world line spacing sampled data approach used with LTE in cref{ch:paper2} are adopted for the RDN and ESRGAN models.
Cref{fig:modcomp} shows RDN and ESRGAN in predicting accurate features when upscaling \qty{320}{\m} data to \qty{80}{\m} line spacing.
It can be inferred that the challenge of super-resolution for these tasks is predominantly a question of providing suitable and sufficient training data, and is less constrained to neural network choices.

\begin{figure}
    \includegraphics[width=\linewidth]{fig/etc/TMI Comparison_23.pdf} % TODO Update for RDN, ESRGAN, LTE, ~3 hours ea.
    \caption{
        A comparison between all deep learning super-resolution methods presented in cref{paper1} and cref{paper2}, using the realistic low-resolution transform developed in cref{paper2}.
        a) The low-resolution input, b) The high-resolution target, c) The RDN+ Network, d) The ESRGAN+ network, and e) The LTE network.}
    \label{fig:modcomp}
\end{figure}

@@

\section{Image quality assessment for potential field grids}
Model performance is quantified using image quality assessments (IQA) such as peak signal-to-noise ratio (PSNR), structural similarity index (SSIM), feature similarity index (FSIM), or many others.
For supervised tasks, a full reference metric can be calculated between the super-resolution prediction and the high-resolution target.
In the Cref{ch:paper1,ch:paper2} works, PSNR and SSIM are used to quantify the performance of the proposed methods in transforming low-resolution grids to high resolution.
However, properties of magnetic grid data are distinct to the images for which these metrics are designed.
%   While SSIM uses structural information present in the Fourier transformed input in combination with local gradient magnitude for value accuracy,
Consistently high SSIM scores are assigned to grids that are perceptually poor matches to the target.
High values are noted in other geophysical grid investigations reporting SSIM metrics \parencite{wangDeeplearningbasedSeismicData2018,bavandsavadkoohiHighresolutionAeromagneticMap2023}
An investigation to identify a more appropriate IQA for potential field grids is required, which accounts for the distinct nature of these data and reflects the predicted quality of geological interpretation.
This may be benefit by automated analysis methods described above.

\section{Physics informed neural networks}
Potential fields are constrained by several well understood physical laws, and gridded data ideally conform to these properties for use in inversion applications and further numerical processing.
Geological interpretation is less reliant on this adherence, however \emph{a priori} information used in physics informed neural networks improve training speed and model performance \parencite{raissiPhysicsinformedNeuralNetworks2019}, and are especially useful in data-constrained domains such as geophysics.
It is possible to broaden the definition and extend the concept of informed networks to the use of generative adversarial networks in Cref{paper1}.
In these networks, the informing method is termed discriminator loss, and is performed by a classification network conditioned to identify images as `realistic' or `not realistic', depending on features learnt to be important within the target data by the discriminator.
Using crafted or learnt criterions to inform a neural network greatly assists with learning high quality models, and either approach may be useful for to geophysics in addressing the paucity of ground truth.
In addition, demonstrating the enforcement of physical properties of potential fields for model outputs may address apprehension for machine learning in geophysics practice.

@@

\chapter{Conclusions}
This thesis presented contemporary methods in deep learning to enhance potential field methods in geoscience.
By including both synthetic and real survey data for training and case studies, it was established that these methods could generalise to real-world applications.

To the best of my knowledge, Cref{ch:paper1} presents the first investigation in the literature of super-resolution enhancement of gridded aerogeophysical data.
Prior to this work, it was unknown how well SR methods developed for natural images in computer vision research would adapt to the features present within magnetic grids, or the challenges faced in developing a training dataset for these data.
We established that the CNN SR methods RDN and ESRGAN could be trained to upsample low-resolution textures in magnetic grids by four times scale, and that the results were more accurate than bicubic upsampling in the real-world case study region.
The GAN method used in Cref{ch:paper2} enhanced resolution, however it introduced significant spurious features, masking the enhancement.
Hence, the method based on RDN was preferred.
We identified the need for sufficient representative training data, and a realistic method to generate low-resolution training data.

The challenge of training data availability was addressed in Cref{ch:paper2}, where we investigated the use of geologically varied synthetic data to train a baseline model, fine-tuning with real-world data, and simulating realistic line sampling to create low-resolution data.
We implemented the contemporary LTE network to perform this investigation, these data sampled at four times wider line spacing showed increased structural accuracy when fine-tuned with real-world data.
Additionally, spurious features in the baseline model were correctly resolved when fine-tuned.
We identified opportunities for more detailed investigations using a filtered approach in selecting training data, and these were outlined in \Cref{ch:futurework}.

The contribution in Cref{ch:paper3} used coordinate multilayer perceptron neural networks to learn a representation of the function that describes a surveyed potential field extent.
To my knowledge, this is the first time implicit neural representation has been reported for use with aerogeophysical survey data in the literature.
These networks are highly suited for processing spatial data such as point sampled geophysics surveys.
The contribution is capable of predicting regularly spaced grids from scattered survey point data, and a novel method using the representation to calculate derivatives of a potential field, both fundamental to interpretation and processing in geophysics.
By commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
Coordinate MLPs for implicit representation are a currently burgeoning field of research in computer science, and we identify opportunities in extending advances in this research to future applications in geophysics, as outlined in \Cref{ch:futurework}.

>>>>

Super-resolution for geophysics is still in its infancy and requires further efforts towards improving reliability before wide-spread adoption in practice.
This section proposes four avenues for future research:
\begin{enumerate}
    \item{} Exploring the outcomes from the use of SR in the task of geological mapping
    \item{} Understanding the impact of choices in constructing a model and dataset for SR geophysics for a range of geological domains,
    \item{} Identifying an improved image quality assessment for gridded potential field data.
    \item{} Investigating the use of a physics-informed approach for SR and implicit representation, and,
\end{enumerate}

\section{SR impact on interpretation}
The super-resolution outputs from the methods presented in this thesis have improved structural accuracy over the low-resolution grids.
This is likely to have a positive impact on the detection of features such as lineaments and unit boundaries for geological interpretation from magnetic data.
However, despite the reduction of introduced structural inaccuracies with the use of open access training data in \Cref{ch:paper2}, it is likely that some features with specific structural geometries will still be incorrectly super-resolved.
A domain expert may be able to recognise structures that are inconsistent using auxiliary information and the regional geological context, but machine processed super-resolution grids will use these data as provided.
In order to improve the adoption of SR for geological interpretation from potential field data, a rigorous study of the impact of SR on interpretation outcomes is required.
This could be a geological mapping task, such as the reinterpretation of regional geology using super-resolution upscaled regional magnetic TMI grids.
Comparison could then be made for the accuracy and benefit of a super-resolution method compared to the existing low-resolution interpretation.
However, such analysis will be difficult to perform with respect to different interpreters' variability in pattern recognition, especially in the context of their confidence in the reliability of the super-resolution product.
Previously, \textcite{sivarajahIdentifyingEffectiveInterpretation2013} reported significant variability between individuals when detecting geological targets from magnetic geophysics data.
A more reliable, repeatable and quantifiable approach would use automated pattern recognition methods developed for geophysical features.
For example, the automated lineament detection technique of \textcite{holdenAutomaticIdentificationResponses2011,holdenAutomatedAnalysisRegional2008} could be applied to low-resolution and SR grids, to compare their delineated features.

\section{SR model construction}
Several models were implemented throughout \Cref{ch:paper1,ch:paper2}, namely RDN, ESRGAN+, and LTE\@.
% However, the most important choice in developing a performant super-resolution neural network for real-world aeromagnetic survey data was the choice of training data.
\Cref{ch:paper1} was prepared prior the release of a large and varied synthetic geology training dataset \parencite{jessellNoddyverseMassiveData2022}.
It was seen in \Cref{ch:paper2} that this data improved model performance for a range of features by using transfer learning for pre-training.
% This transfer learning approach for training data is also found to be useful for RDN and ESRGAN\@.
Here, the use of synthetic and real-world line spacing sampled data approach used with LTE in \Cref{ch:paper2} are adopted for the RDN+ and ESRGAN+ models.
\Cref{fig:modcomp} shows RDN and ESRGAN+ in predicting accurate features when upscaling \qty{320}{\m} data to \qty{80}{\m} line spacing.
It can be inferred that the challenge of super-resolution for these tasks is predominantly a question of providing suitable and sufficient training data, and is less constrained to neural network choices.

\begin{figure}
    \includegraphics[width=\linewidth]{fig/etc/TMI Comparison_23.pdf} % TODO Update for RDN, ESRGAN, LTE, ~3 hours ea.
    \caption{
        A comparison between all deep learning super-resolution methods presented in \Cref{ch:paper1} and \Cref{ch:paper2}, using the realistic low-resolution transform developed in \Cref{ch:paper2}.
        a) The low-resolution input, b) The high-resolution target, c) The RDN+ Network, d) The ESRGAN+ network, and e) The LTE network.}
    \label{fig:modcomp}
\end{figure}

@@

\section{Image quality assessment for potential field grids}
Model performance is quantified using image quality assessments (IQA) such as peak signal-to-noise ratio (PSNR), structural similarity index (SSIM), feature similarity index (FSIM), or many others.
For supervised tasks, a full reference metric can be calculated between the super-resolution prediction and the high-resolution target.
In the \Cref{ch:paper1,ch:paper2} works, PSNR and SSIM are used to quantify the performance of the proposed methods in transforming low-resolution grids to high resolution.
However, properties of magnetic grid data are distinct to the images for which these metrics are designed.
%   While SSIM uses structural information present in the Fourier transformed input in combination with local gradient magnitude for value accuracy,
Consistently high SSIM scores are assigned to grids that are perceptually poor matches to the target.
High values are noted in other geophysical grid investigations reporting SSIM metrics \parencite{wangDeeplearningbasedSeismicData2018,bavandsavadkoohiHighresolutionAeromagneticMap2023}.
An investigation to identify a more appropriate IQA for potential field grids is required, which accounts for the distinct nature of these data and reflects the predicted quality of geological interpretation.
This may be benefit by automated analysis methods described above.

\section{Physics-informed neural networks}
Potential fields are constrained by several well understood physical laws, and gridded data ideally conform to these properties for use in inversion applications and further numerical processing.
Geological interpretation is less reliant on this adherence, however \emph{a priori} information used in physics-informed neural networks improve training speed and model performance \parencite{raissiPhysicsinformedNeuralNetworks2019}, and are especially useful in data-constrained domains such as geophysics.
It is possible to broaden the definition and extend the concept of informed networks to the use of generative adversarial networks in \Cref{ch:paper1}.
In these networks, the informing method is termed discriminator loss, and is performed by a classification network conditioned to identify images as `realistic' or `not realistic', depending on features learnt to be important within the target data by the discriminator.
Using crafted or learnt criterions to inform a neural network greatly assists with learning high quality models, and either approach may be useful for to geophysics in addressing the paucity of ground truth.
In addition, demonstrating the enforcement of physical properties of potential fields for model outputs may address apprehension for machine learning in geophysics practice.

@@

\chapter{Conclusions}
\label{ch:conclusions}
This thesis presented contemporary methods in deep learning to enhance potential field methods in geoscience.
By including both synthetic and real survey data for training and case studies, it was established that these methods could generalise to real-world applications.

To the best of my knowledge, \Cref{ch:paper1} presents the first investigation in the literature of super-resolution enhancement of gridded aerogeophysical data.
Prior to this work, it was unknown how well SR methods developed for natural images in computer vision research would adapt to the features present within magnetic grids, or the challenges faced in developing a training dataset for these data.
We established that the CNN SR methods RDN and ESRGAN could be trained to upsample low-resolution textures in magnetic grids by four times scale, and that the results were more accurate than bicubic upsampling in the real-world case study region.
The GAN method used in \Cref{ch:paper2} enhanced resolution, however it introduced significant spurious features, masking the enhancement.
Hence, the method based on RDN was preferred.
We identified the need for sufficient representative training data, and a realistic method to generate low-resolution training data.

The challenge of training data availability was addressed in \Cref{ch:paper2}, where we investigated the use of geologically varied synthetic data to train a baseline model, fine-tuning with real-world data, and simulating realistic line sampling to create low-resolution data.
We implemented the contemporary LTE network to perform this investigation, these data sampled at four times wider line spacing showed increased structural accuracy when fine-tuned with real-world data.
Additionally, spurious features in the baseline model were correctly resolved when fine-tuned.
We identified opportunities for more detailed investigations using a filtered approach in selecting training data, and these were outlined in \Cref{ch:futurework}.

The contribution in \Cref{ch:paper3} used coordinate multilayer perceptron neural networks to learn a representation of the function that describes a surveyed potential field extent.
To my knowledge, this is the first time implicit neural representation has been reported for use with aerogeophysical survey data in the literature.
These networks are highly suited for processing spatial data such as point sampled geophysics surveys.
The contribution is capable of predicting regularly spaced grids from scattered survey point data, and a novel method using the representation to calculate derivatives of a potential field, both fundamental to interpretation and processing in geophysics.
By commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
Coordinate MLPs for implicit representation are a currently burgeoning field of research in computer science, and we identify opportunities in extending advances in this research to future applications in geophysics, as outlined in \Cref{ch:futurework}.

### 2023-10-10 intro.tex
Further fix cross references

\section{Research Contributions and Thesis Structure}
This thesis is presented as a series of papers which are published or in-preparation for publication.
These are presented in Cref{ch:paper1,ch:paper2,ch:paper3}, each with a brief contextual introduction.
Key findings and contributions of the research are discussed in Cref{ch:discussion}, followed by overall thesis conclusions in Cref{ch:conclusion}.

Cref{ch:paper1} presents published work on an initial investigation into the technique of deep learning super-resolution for magnetic potential field data.
To our knowledge, this chapter presents the first investigation of deep learning super-resolution of gridded magnetic potential field data within the literature.
Two convolutional neural network models namely RDN and ESRGAN from the contemporary literature are implemented \parencite{zhangResidualDenseNetwork2018,limEnhancedDeepResidual2017}.
With uncertainty on the amount of data required to train a SR model for magnetic grids, a dataset comprising readily available high- and low-resolution TMI grid pairs is formed using published state magnetic compilation maps of Western Australia \parencite{brett20MagneticMerged2020}.
These data contain magnetic features and textures from a range of geological domains.
The chapter concludes by observing the success of SR on magnetic features, characterising the performance of the two adapted networks, and recommending the method based on RDN as being more reliable for the SR task.
The primary contribution of Cref{ch:paper1} is establishing the capacity of the SR networks developed for natural images in computer vision to extend to the distinctly different data of low-resolution aeromagnetic geophysics.

Following the findings of the previous chapter, Cref{ch:paper2} addresses the need for more generalisable model training for real-world super-resolution.
It undertakes these aims by training a baseline synthetic model with data from a contemporaneously published collection of magnetic grids forward modelled from realistic geology \parencite{jessellNoddyverseMassiveData2022}, on data from over \num{300} distinct geological histories.
It subsequently fine-tune trains the baseline model on real-world high-resolution state magnetic data.
Furthermore, it adopts advances in the machine learning literature by implementing the LTE network \parencite{leeLocalTextureEstimator2022}.
The proposed method simulates aeromagnetic line sampling of a potential field extent at \qty{320}{\m} and \qty{80}{\m}, which are in the regional and prospect scale respectively.
While the baseline model can generalise to real-world data, some incorrect reconstructions remain.
By fine-tune training the model with transformed real-world state map data, these errors are rectified, and the structural accuracy is improved.

Cref{ch:paper3} reports the use of coordinate multilayer perceptron (CMLP) neural networks for representation learning of a surveyed potential field extent.
With the proposed method, scattered line data can be encoded into the grid, and cells queried at interpolating coordinates to predict regularised grids.
CMLP networks are highly suited for processing spatial data such as point sampled geophysics surveys.
Additionally, spatial derivatives are calculated on the learnt function directly.
%TODO ensure refer back to aims
Both tasks are fundamental to interpretation and processing in geophysics, and by commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
To our knowledge, this research is the first description of deep learning implicit neural representation for aeromagnetic potential field data in the literature.

Cref{ch:discussion} presents a synthesis discussion for the prior chapters. Given the recency of deep learning super-resolution and implicit neural representation, multiple avenues of future opportunities are identified.
These aim to increase the adoption of super-resolution in geophysics by addressing apprehension toward the reliability of interpretation of the outputs.
A synthesis comparison of the SR networks adapted in Cref{ch:paper1,ch:paper2} is also presented, which prompts questions regarding model and dataset construction.
The use of a priori information well understood to geophysics is also discussed, with potential application in both super-resolution and implicit neural representation.

Finally, overall conclusions are presented in Cref{ch:conclusion}.

>>>>

\section{Research Contributions and Thesis Structure}
This thesis is presented as a series of papers which are published or in-preparation for publication.
These are presented in Cref{ch:paper1,ch:paper2,ch:paper3}, each with a brief contextual introduction.
Key findings and summary conclusions of the research are discussed in \Cref{ch:conclusions}.

Cref{ch:paper1} presents published work on an initial investigation into the technique of deep learning super-resolution for magnetic potential field data.
To our knowledge, this chapter presents the first investigation of deep learning super-resolution of gridded magnetic potential field data within the literature.
Two convolutional neural network models namely RDN and ESRGAN from the contemporary literature are implemented \parencite{zhangResidualDenseNetwork2018,limEnhancedDeepResidual2017}.
With uncertainty on the amount of data required to train a SR model for magnetic grids, a dataset comprising readily available high- and low-resolution TMI grid pairs is formed using published state magnetic compilation maps of Western Australia \parencite{brett20MagneticMerged2020}.
These data contain magnetic features and textures from a range of geological domains.
The chapter concludes by observing the success of SR on magnetic features, characterising the performance of the two adapted networks, and recommending the method based on RDN as being more reliable for the SR task.
The primary contribution of \Cref{ch:paper1} is establishing the capacity of the SR networks developed for natural images in computer vision to extend to the distinctly different data of low-resolution aeromagnetic geophysics.

Following the findings of the previous chapter, \Cref{ch:paper2} addresses the need for more generalisable model training for real-world super-resolution.
It undertakes these aims by training a baseline synthetic model with data from a contemporaneously published collection of magnetic grids forward modelled from realistic geology \parencite{jessellNoddyverseMassiveData2022}, on data from over \num{300} distinct geological histories.
It subsequently fine-tune trains the baseline model on real-world high-resolution state magnetic data.
Furthermore, it adopts advances in the machine learning literature by implementing the LTE network \parencite{leeLocalTextureEstimator2022}.
The proposed method simulates aeromagnetic line sampling of a potential field extent at \qty{320}{\m} and \qty{80}{\m}, which are in the regional and prospect scale respectively.
While the baseline model can generalise to real-world data, some incorrect reconstructions remain.
By fine-tune training the model with transformed real-world state map data, these errors are rectified, and the structural accuracy is improved.

\Cref{ch:paper3} reports the use of coordinate multilayer perceptron (CMLP) neural networks for representation learning of a surveyed potential field extent.
With the proposed method, scattered line data can be encoded into the grid, and cells queried at interpolating coordinates to predict regularised grids.
CMLP networks are highly suited for processing spatial data such as point sampled geophysics surveys.
Additionally, spatial derivatives are calculated on the learnt function directly.
%TODO ensure refer back to aims
Both tasks are fundamental to interpretation and processing in geophysics, and by commencing the methods from point sample data, the number of individual processing steps is reduced, and the application is fully trainable.
To our knowledge, this research is the first description of deep learning implicit neural representation for aeromagnetic potential field data in the literature.

\Cref{ch:futurework} presents a synthesis discussion for the prior chapters. Given the recency of deep learning super-resolution and implicit neural representation, multiple avenues of future opportunities are identified.
These aim to increase the adoption of super-resolution in geophysics by addressing apprehension toward the reliability of interpretation of the outputs.
A synthesis comparison of the SR networks adapted in \Cref{ch:paper1,ch:paper2} is also presented, which prompts questions regarding model and dataset construction.
The use of \emph{a priori} information well understood to geophysics is also discussed, with potential application in both super-resolution and implicit neural representation.

Finally, overall conclusions are presented in \Cref{ch:conclusions}.
